Bayesian experimental design using regularized \\ determinantal point processes

### 2022-11-23 design/aistats/bayesian.tex
Updates from Overleaf
Working compile

%   \usepackage{natbib}
%   \renewcommand{\bibname}{References}
%   \renewcommand{\bibsection}{\subsubsection*{\bibname}}
% \fi
\documentclass[thesis.tex]{subfiles}

@@

% \begin{abstract}
We establish a fundamental connection between Bayesian
experimental design and determinantal point processes
(DPPs). Experimental design is a classical task in combinatorial
optimization, where we wish to select a small subset of $d$-dimensional
 vectors to minimize a statistical optimality criterion.
 We show that a new regularized variant of
 DPPs can be used to design
efficient algorithms for finding $(1+\epsilon)$-approximate solutions
to experimental design under four commonly used optimality
criteria: A-, C-, D- and V-optimality. A key novelty is that we offer
improved guarantees under the Bayesian framework.
Our algorithm returns a $(1+\epsilon)$-approximate solution when the
subset size $k$ is
$\Omega(\frac{d_{\A}}{\epsilon} + \frac{\log
1/\epsilon}{\epsilon^2})$, where $d_{\A}\ll d$ is an effective dimension
determined by prior knowledge (via a precision matrix $\A$). This is the first
approximation guarantee where the dependence on $d$ is
replaced by an effective dimension. Moreover, the time complexity
of our algorithm significantly improves on existing approaches
with comparable guarantees. 
% \end{abstract}

@@

Consider a collection of $n$ experiments parameterized by
$d$-dimensional vectors $\x_1,\dots,\x_n$, and let $\X$ denote the
$n\times d$ matrix with rows $\x_i^\top$. The outcome of the $i$th
experiment is a random variable $y_i = \x_i^\top\w + \xi_i$, where
$\w$ is the parameter vector of a linear model with prior distribution
$\Nc(\zero,\sigma^2\A^{-1})$, and $\xi_i\sim \Nc(0,\sigma^2)$ is
independent noise. In experimental design, we have access to the vectors 
$\x_i^\top$, for $i\in\{1,\ldots,n\} = [n]$, but we are allowed to
observe only a small number of outcomes $y_i$ for experiments we choose.
Suppose that we observe the outcomes from a subset
$S\subseteq [n]$ of $\lvert S \vert = k$ experiments. The
posterior distribution of $\w$ given $\y_S$ (the vector of outcomes in $S$) is:
\begin{align*}
  \w\mid \y_S \ &\sim\ \Nc(\mub, \Sigmab), \\
  \text{where } \mub &= (\X_S^\top\X_S + \A)^{-1}\X_S^\top\y_S, \\
  \Sigmab &= \sigma^2(\X_S^\top\X_S+\A)^{-1}.
\end{align*}
Here, $\X_S$ is the $k\times d$ matrix with rows $\x_i^\top$ for
$i\in S$.

@@

\textbf{Question 1:} \quad
Given $\X$, $f_{\A}$ and $k$, can we efficiently find a $(1+\epsilon)$-approximation for
$\opt$?

\textbf{Question 2:} \quad
Given only $\Sigmab_{\X}$, $f_{\A}$ and $k$,
what is the upper bound on $\opt$?

@@

  \paragraph{Answer to Question 1.}
We propose an efficient $(1+\epsilon)$-approximation algorithm for
A/C/D/V-optimal Bayesian  experimental design:
\begin{theorem}\label{t:q2}
  Let $f_{\A}$ be A/C/D/V-optimality and $\X$ be $n\times d$. If
  $k=\Omega\big(\frac{d_{\A}}{\epsilon} +
  \frac{\log1/\epsilon}{\epsilon^2}\big)$ for some $\epsilon\in(0,1)$, then 
we can find in polynomial time a subset $S$ of size $k$ s.t.
  \begin{align*}
    f_{\A}\big(\X_S^\top\X_S\big) \leq (1+\epsilon)\cdot\opt.
  \end{align*}
\end{theorem}
\begin{remark}
  The algorithm referred to in Theorem~\ref{t:q2} first solves a convex
  relaxation of the task via a semi-definite program (SDP) to find a
  weight vector $p\in[0,1]^n$, then uses our new randomized algorithm 
  to round the weights to $\{0,1\}$, obtaining the subset
  $S$. The expected cost after SDP is $O(ndk+k^2d^2)$.
\end{remark}
% Note that, unlike in Theorem \ref{t:q1}, we use the unrescaled effective
% dimension instead of the rescaled one, $d_{\frac nk\!\A}$. The
% actual effective dimension that applies here (given in the proof in
% Section \ref{s:guarantees}) depends on the SDP solution. It is
% always upper bounded by $d_{\A}$, but it may be significantly
% smaller.

A number of recent works studied $(1+\epsilon)$-approximate SDP-based algorithms 
for classical and Bayesian experimental design (see Table
\ref{tab:comp} and Section \ref{s:related-work} for a comparison). Unlike \emph{all} prior work on
this topic, we are able to eliminate the
dependence of the subset size $k$ on the dimension $d$, replacing it
with the potentially much smaller effective dimension $d_\A$.
Our result also improves over the existing approaches in terms of the computational
cost of the rounding procedure that is performed after solving the SDP. A
number of different methods can be used to solve the SDP relaxation (see
Section \ref{s:experiments}). For
example, \cite{near-optimal-design} suggest using an iterative
optimizer called entropic mirror descent, which is known to exhibit
fast convergence and can run in $O(nd^2T)$ time, where $T$ is the
number of iterations.

\begin{figure*}[htbp]
  \vspace{-2mm}
  \centering
  \renewcommand{\arraystretch}{1.5}
\begin{tabular}{r||c|c|l|l}
  &Criteria&Bayesian& $k$ %& dep.~on $d$ & dep.~on $1/\epsilon$
  & Cost after SDP\\%of rounding\\
  \hline\hline
\small  \cite{tractable-experimental-design}
  &\small A,V&\xmark  &$d^2/\epsilon$
  % &\Red{quadratic}& \Orange{linear}
  &$n^2\cdot d$\\
\small  \cite{near-optimal-design}
  &\small A,C,D,E,G,V&\cmark
&$ d/\epsilon^2$
% &\Orange{linear}& \Red{quadratic}
  &$n\cdot kd^2$\\ 
\small  \cite{proportional-volume-sampling}
  &\small A,D&\xmark &$d/\epsilon$
  % & \Orange{linear}& \Orange{linear}
  & $n^4\cdot k^2d$\\ 
  \hline
  \small\textbf{this paper}&\small A,C,D,V&\cmark
    & $d_{\A}/\epsilon$
    % &\Green{none}& \Orange{linear}
  & $n\cdot kd + k^2d^2$
\end{tabular}
%\vspace{2mm}
\captionof{table}{Comparison of SDP-based $(1+\epsilon)$-approximation algorithms for
  classical and Bayesian experimental design (X-mark means that only
  the classical setting applies).
In the cost analysis, $n$ could be replaced by the number of
non-zero weights in the SDP solution. For simplicity we omit the log
terms and assume that $\epsilon = \Omega(\frac1{d_{\A}})$. Our
approach beats other methods both in terms of the
runtime and the dependence of $k$ on $d$ (when $d_\A=o(d)$).} 
\label{tab:comp}
\vspace{-3mm}
\end{figure*}

\paragraph{Answer to Question 2.}
%Formalizing the above intuitions,
By performing a careful theoretical analysis of the performance of our
algorithm, we are able to give an improved upper bound on $\opt$. In the below
result, we use a more refined notion of effective dimensionality for
Bayesian experimental design,
$d_{\frac nk\A}$ (where the precision matrix $\A$ is scaled by factor $\frac
nk$), which is smaller than $d_\A$ and therefore leads 
to a tighter bound.
% $d_{\frac nk\A}\leq d_{\A}$ as the notion of 
% effective dimensionality, which is an improvement over the prior work
% We now present our first main result, which provides an analytic bound
% on the four optimality criteria listed above (Question 1) for most
% practically relevant values of $k$.
\begin{theorem}\label{t:q1}
  Let $f_{\A}$ be A/C/D/V-optimality and $\X$ be
  $n\times d$. For any $k$ such that $k\geq 4d_{\frac
    nk\!\A}$, %we have
  \begin{align*}
\opt    \leq \bigg(\ 1+8\,\frac{d_{\frac
    nk\!\A}}{k} + 8\sqrtshort{\frac{\ln (k/d_{\frac nk\!\A})}{k}}\
    \bigg)\cdot f_{\A}\big(\tfrac kn\Sigmab_{\X}\big).
  \end{align*}
\end{theorem}
\begin{remark}
  We give a (randomized) algorithm which (with probability 1) finds
  the subset $S$ that certifies this bound and has expected time
  complexity $O(ndk+k^2d^2)$. 
  % and is $O(t (n d^2 + k d^2))$ time with
  % probability exponentially increasing in $t$.
  % Michal: please check this statement.
\end{remark}
  In particular, this means that if $k\geq 4d_{\frac nk\!\A}$
  then there is $S$ of size $k$
  which satisfies $f_{\A}(\X_S^\top\X_S) = O(1)\cdot f_{\A}(\frac kn\Sigmab_{\X})$.
  This not only improves on \cite{regularized-volume-sampling} in terms
  of the supported range of sizes $k$, but also in terms of the obtained bound (see
  Section \ref{s:related-work} for a~comparison). In Section
  \ref{s:experiments}, we we provide numerical evidence suggesting
  that for many real datasets the quantity $f_{\A}(\frac
  kn\Sigmab_{\X})$ provides a good estimate for $\opt$ to within a
  factor of 2.

Theorem \ref{t:q1} suggests that the right notion of degrees of
freedom for Bayesian experimental design can in fact be smaller
than $d_{\A}$.  
Intuitively, since $d_{\A}$ is computed using the full data covariance
$\Sigmab_{\X}$, it is not in the same scale as the smaller covariance
$\X_S^\top\X_S$ based on the subset $S$ of size $k\ll n$. In our
result this is corrected by increasing the 
regularization on $\Sigmab_\X$ from $\A$ to $\frac{n}{k}\A$ and using
$d_{\frac nk\A}=d_{\frac{n}{k}\A}(\Sigmab_\X)$ as the degrees of
freedom. 
Note that $d_{\frac nk\!\A}\leq d_{\A}$ and this gap can be very large for some
problems (see discussion in Appendix~\ref{a:deff}).

@@

  Given a psd matrix $\A$ and a weight vector $p=(p_1,...,p_n)\in[0,1]^n$, we
define $\DPPreg{p}(\X,\A)$ as a distribution over subsets $S\subseteq[n]$ (of all
sizes) such that (see Definition~\ref{d:r-dpp}):
\begin{align*}
 \Pr(S)\propto \det(
  \X_S^\top\X_S+\A)\ \cdot \prod_{i\in S}p_{i}\cdot\prod_{i\not\in
  S}(1-p_i).
\end{align*}
A number of regularized DPPs have been proposed recently
\citep{dpp-intermediate,regularized-volume-sampling}, mostly within the context
of Randomized Numerical Linear Algebra (RandNLA)~
\citep{Mah-mat-rev_JRNL,DM16_CACM,RandNLA_PCMIchapter_TR}.
To our knowledge, ours is the first such definition that strictly falls
under the umbrella of traditional DPPs \citep{dpp-ml}.
We show this in Section \ref{s:r-dpp}, where we also prove that
regularized DPPs can be decomposed into a low-rank DPP plus
i.i.d. Bernoulli sampling (Theorem \ref{t:algorithm}). This decomposition reduces
the sampling cost from $O(n^3)$ to $O(nd^2)$,
and involves a more general result about DPPs defined via a
correlation kernel (Lemma~\ref{l:decomposition}), which is of
independent interest.

@@

Purely greedy approximation algorithms have been shown to provide
guarantees in a number of special cases for experimental design. One
example is classical D-optimality criterion, which can be converted to
a submodular function \citep{submodularity-optimal-design}. Also, greedy algorithms
for Bayesian A/V-optimality criteria have been considered by
\cite{greedy-supermodular} and \cite{greedy-graph-sampling}. These
methods can only provide a constant factor approximation guarantee
(as opposed to $1+\epsilon$), and the factor is generally problem
dependent (which means it could be arbitrarily large).
Finally, a number of
heuristics with good empirical performance have been proposed, such as Fedorov's
exchange method \citep{cook1980comparison}. However, in this work
we focus on methods that provide theoretical approximation guarantees.
% In Section \ref{s:experiments} we experimentally show that
% our regularized DPP sampling (without solving an SDP) performs
% better than iid sampling methods like uniform and predictive length
% sampling \cite{zhu2015optimal} and that by using the SDP solution
% we can improve it to match the performance of
% greedy algorithms for A-optimal design \cite{chamon2017approximate,greedy-supermodular}
% which empirically have been found to perform well on real world data.

\section{A new regularized determinantal point process}
\label{s:r-dpp}
  In this section we develop the theory for a novel regularized extension of
  determinantal point processes (DPP) which we use as the sampling distribution
  for obtaining guarantees in Bayesian experimental design.
  DPPs form a family of distributions which are used to model repulsion between
  elements in a random set, with many applications in machine learning
  \citep{dpp-ml}. Here, we focus on the setting where we are sampling out of all
  $2^n$ subsets $S\subseteq[n]$. Traditionally, a DPP is defined by a
  correlation kernel, which is an $n\times n$ psd matrix $\K$ with eigenvalues
  between 0 and 1, i.e., such that $\zero\preceq\K\preceq \I$. Given a
  correlation kernel $\K$, the corresponding DPP is
  defined as
  \begin{align*}
    S\sim\DPPcor(\K)\quad\text{iff}\quad
    \Pr(T\subseteq S) = \det(\K_{T,T})\ \  \forall_{T\in[n]},
  \end{align*}
where $\K_{T,T}$ is the submatrix of $\K$ with rows and columns
indexed by $T$. Another way of defining a DPP, popular in the machine
learning community, is via an ensemble kernel $\L$. Any psd matrix
$\L$ is an ensemble kernel of a DPP defined as:
  \begin{align*}
    S\sim\DPPens(\L)\quad\text{iff}\quad \Pr(S) \propto \det(\L_{S,S}).\quad~
  \end{align*}
Crucially, every $\DPPens$ is also a $\DPPcor$, but not the other way
around. Specifically, $\DPPens(\L)=\DPPcor(\K)$ when:
\begin{align*}
  &\text{(a)}\ \  \L=\K(\I-\K)^{-1},\qquad
  &\text{(b)}\ \  \K = \I -  (\I+\L)^{-1},
\end{align*}
but (a) requires that $\I-\K$ be invertible which is not true
for some DPPs.
(This will be important in our analysis.)
The classical algorithm for sampling from a DPP requires the
eigendecomposition of either matrix $\K$ or $\L$, which in general
costs $O(n^3)$, followed by a sampling procedure which costs
$O(n\,|S|^2)$ \cite{dpp-independence,dpp-ml}.

>>>>

%   \usepackage{natbib}
%   \renewcommand{\bibname}{References}
%   \renewcommand{\bibsection}{\subsubsection*{\bibname}}
% \fi
\documentclass[../../thesis.tex]{subfiles}

@@

% \begin{abstract}

\chapter{Bayesian experimental design with regularized determinantal point processes}
\label{ch:design}

In this chapter, we establish a fundamental connection between Bayesian
experimental design and determinantal point processes
(DPPs). Experimental design is a classical task in combinatorial
optimization, where we wish to select a small subset of $d$-dimensional
vectors to minimize a statistical optimality criterion.
We show that a new regularized variant of
DPPs can be used to design
efficient algorithms for finding $(1+\epsilon)$-approximate solutions
to experimental design under four commonly used optimality
criteria: A-, C-, D- and V-optimality. A key novelty is that we offer
improved guarantees under the Bayesian framework.
Our algorithm returns a $(1+\epsilon)$-approximate solution when the
subset size $k$ is
$\Omega(\frac{d_{\A}}{\epsilon} + \frac{\log
    1/\epsilon}{\epsilon^2})$, where $d_{\A}\ll d$ is an effective dimension
determined by prior knowledge (via a precision matrix $\A$). This is the first
approximation guarantee where the dependence on $d$ is
replaced by an effective dimension. Moreover, the time complexity
of our algorithm significantly improves on existing approaches
with comparable guarantees. Some of the results here were initially
published in \fullcite{bayesian-experimental-design}.
% \end{abstract}

@@

Consider a collection of $n$ experiments parameterized by
$d$-dimensional vectors $\x_1,\dots,\x_n$, and let $\X$ denote the
$n\times d$ matrix with rows $\x_i^\top$. The outcome of the $i$th
experiment is a random variable $y_i = \x_i^\top\w + \xi_i$, where
$\w$ is the parameter vector of a linear model with prior distribution
$\Nc(\zero,\sigma^2\A^{-1})$, and $\xi_i\sim \Nc(0,\sigma^2)$ is
independent noise. In experimental design, we have access to the vectors
$\x_i^\top$, for $i\in\{1,\ldots,n\} = [n]$, but we are allowed to
observe only a small number of outcomes $y_i$ for experiments we choose.
Suppose that we observe the outcomes from a subset
$S\subseteq [n]$ of $\lvert S \vert = k$ experiments. The
posterior distribution of $\w$ given $\y_S$ (the vector of outcomes in $S$) is:
\begin{align*}
  \w\mid \y_S \      & \sim\ \Nc(\mub, \Sigmab),                 \\
  \text{where } \mub & = (\X_S^\top\X_S + \A)^{-1}\X_S^\top\y_S, \\
  \Sigmab            & = \sigma^2(\X_S^\top\X_S+\A)^{-1}.
\end{align*}
Here, $\X_S$ is the $k\times d$ matrix with rows $\x_i^\top$ for
$i\in S$.

@@

\begin{question}\label{q:bayesian:q1}
  Given $\X$, $f_{\A}$ and $k$, can we efficiently find a $(1+\epsilon)$-approximation for
  $\opt$?
\end{question}

\begin{question}\label{q:bayesian:q2}
  Given only $\Sigmab_{\X}$, $f_{\A}$ and $k$,
  what is the upper bound on $\opt$?
\end{question}

@@

\paragraph{Answer to \Cref{q:bayesian:q1}}
We propose an efficient $(1+\epsilon)$-approximation algorithm for
A/C/D/V-optimal Bayesian  experimental design:
\begin{theorem}\label{t:q2}
  Let $f_{\A}$ be A/C/D/V-optimality and $\X$ be $n\times d$. If
  $k=\Omega\big(\frac{d_{\A}}{\epsilon} +
    \frac{\log1/\epsilon}{\epsilon^2}\big)$ for some $\epsilon\in(0,1)$, then
  we can find in polynomial time a subset $S$ of size $k$ s.t.
  \begin{align*}
    f_{\A}\big(\X_S^\top\X_S\big) \leq (1+\epsilon)\cdot\opt.
  \end{align*}
\end{theorem}
\begin{remark}
  The algorithm referred to in Theorem~\ref{t:q2} first solves a convex
  relaxation of the task via a semi-definite program (SDP) to find a
  weight vector $p\in[0,1]^n$, then uses our new randomized algorithm
  to round the weights to $\{0,1\}$, obtaining the subset
  $S$. The expected cost after SDP is $O(ndk+k^2d^2)$.
\end{remark}
% Note that, unlike in Theorem \ref{t:q1}, we use the unrescaled effective
% dimension instead of the rescaled one, $d_{\frac nk\!\A}$. The
% actual effective dimension that applies here (given in the proof in
% Section \ref{s:guarantees}) depends on the SDP solution. It is
% always upper bounded by $d_{\A}$, but it may be significantly
% smaller.

A number of recent works studied $(1+\epsilon)$-approximate SDP-based algorithms
for classical and Bayesian experimental design (see Table
\ref{tab:comp} and Section \ref{s:related-work} for a comparison). Unlike \emph{all} prior work on
this topic, we are able to eliminate the
dependence of the subset size $k$ on the dimension $d$, replacing it
with the potentially much smaller effective dimension $d_\A$.
Our result also improves over the existing approaches in terms of the computational
cost of the rounding procedure that is performed after solving the SDP. A
number of different methods can be used to solve the SDP relaxation (see
Section \ref{s:bayesian:experiments}). For
example, \cite{near-optimal-design} suggest using an iterative
optimizer called entropic mirror descent, which is known to exhibit
fast convergence and can run in $O(nd^2T)$ time, where $T$ is the
number of iterations.

\begin{figure*}[htbp]
  \centering
  \begin{tabular}{r||c|c|l|l}
                              & Criteria             & Bayesian & $k$            %& dep.~on $d$ & dep.~on $1/\epsilon$
                              & Cost after SDP                                   \\%of rounding\\
    \hline\hline
    \small  \cite{tractable-experimental-design}
                              & \small A,V           & \xmark   & $d^2/\epsilon$
    % &\Red{quadratic}& \Orange{linear}
                              & $n^2\cdot d$                                     \\
    \small  \cite{near-optimal-design}
                              & \small A,C,D,E,G,V   & \cmark
                              & $ d/\epsilon^2$
    % &\Orange{linear}& \Red{quadratic}
                              & $n\cdot kd^2$                                    \\
    \small  \cite{proportional-volume-sampling}
                              & \small A,D           & \xmark   & $d/\epsilon$
    % & \Orange{linear}& \Orange{linear}
                              & $n^4\cdot k^2d$                                  \\
    \hline
    \small\textbf{this paper} & \small A,C,D,V       & \cmark
                              & $d_{\A}/\epsilon$
    % &\Green{none}& \Orange{linear}
                              & $n\cdot kd + k^2d^2$
  \end{tabular}
  \captionof{table}{Comparison of SDP-based $(1+\epsilon)$-approximation algorithms for
    classical and Bayesian experimental design (X-mark means that only
    the classical setting applies).
    In the cost analysis, $n$ could be replaced by the number of
    non-zero weights in the SDP solution. For simplicity we omit the log
    terms and assume that $\epsilon = \Omega(\frac1{d_{\A}})$. Our
    approach beats other methods both in terms of the
    runtime and the dependence of $k$ on $d$ (when $d_\A=o(d)$).}
  \label{tab:comp}
\end{figure*}

\paragraph{Answer to \Cref{q:bayesian:q2}}
%Formalizing the above intuitions,
By performing a careful theoretical analysis of the performance of our
algorithm, we are able to give an improved upper bound on $\opt$. In the below
result, we use a more refined notion of effective dimensionality for
Bayesian experimental design,
$d_{\frac nk\A}$ (where the precision matrix $\A$ is scaled by factor $\frac
  nk$), which is smaller than $d_\A$ and therefore leads
to a tighter bound.
% $d_{\frac nk\A}\leq d_{\A}$ as the notion of 
% effective dimensionality, which is an improvement over the prior work
% We now present our first main result, which provides an analytic bound
% on the four optimality criteria listed above (Question 1) for most
% practically relevant values of $k$.
\begin{theorem}\label{t:q1}
  Let $f_{\A}$ be A/C/D/V-optimality and $\X$ be
  $n\times d$. For any $k$ such that $k\geq 4d_{\frac
        nk\!\A}$, %we have
  \begin{align*}
    \opt    \leq \bigg(\ 1+8\,\frac{d_{\frac
        nk\!\A}}{k} + 8\sqrtshort{\frac{\ln (k/d_{\frac nk\!\A})}{k}}\
    \bigg)\cdot f_{\A}\big(\tfrac kn\Sigmab_{\X}\big).
  \end{align*}
\end{theorem}
\begin{remark}
  We give a (randomized) algorithm which (with probability 1) finds
  the subset $S$ that certifies this bound and has expected time
  complexity $O(ndk+k^2d^2)$.
  % and is $O(t (n d^2 + k d^2))$ time with
  % probability exponentially increasing in $t$.
  % Michal: please check this statement.
\end{remark}
In particular, this means that if $k\geq 4d_{\frac nk\!\A}$
then there is $S$ of size $k$
which satisfies $f_{\A}(\X_S^\top\X_S) = O(1)\cdot f_{\A}(\frac kn\Sigmab_{\X})$.
This not only improves on \cite{regularized-volume-sampling} in terms
of the supported range of sizes $k$, but also in terms of the obtained bound (see
Section \ref{s:related-work} for a~comparison). In Section
\ref{s:bayesian:experiments}, we we provide numerical evidence suggesting
that for many real datasets the quantity $f_{\A}(\frac
  kn\Sigmab_{\X})$ provides a good estimate for $\opt$ to within a
factor of 2.

\subsection{Comparison of different effective dimensions}

Theorem \ref{t:q1} suggests that the right notion of degrees of
freedom for Bayesian experimental design can in fact be smaller
than $d_{\A}$.
Intuitively, since $d_{\A}$ is computed using the full data covariance
$\Sigmab_{\X}$, it is not in the same scale as the smaller covariance
$\X_S^\top\X_S$ based on the subset $S$ of size $k\ll n$. In our
result this is corrected by increasing the
regularization on $\Sigmab_\X$ from $\A$ to $\frac{n}{k}\A$ and using
$d_{\frac nk\A}=d_{\frac{n}{k}\A}(\Sigmab_\X)$ as the degrees of
freedom.
Note that $d_{\frac nk\!\A}\leq d_{\A}$ and this gap can be very large for some
problems.

Consider the two definitions we are comparing:
\begin{description}
  \item[Full effective dimension]\quad\ \ \,$d_{\A} =
    \tr\big(\Sigmab_\X(\A+\Sigmab_\X)^{-1}\big)$,
  \item[Scaled effective dimension] $d_{\frac nk\!\A} =
      \tr\big(\Sigmab_\X(\tfrac nk\A + \Sigmab_\X)^{-1}\big)$.
\end{description}
Here, we demonstrate that these two effective dimensions can be
very different for some matrices and quite similar on others. For
simplicity, we consider two diagonal data covariance matrices as
our examples: \emph{identity covariance}, $\Sigmab_1 = \I$, and an
\emph{approximately low-rank covariance}, $\Sigmab_2 =
  (1-\epsilon)\frac ds \I_S+\epsilon\I$, where $\I_S$ is the
diagonal matrix with ones on the entries indexed by subset
$S\subseteq [d]$ of
size $s<d$ and zeros
everywhere else. The second matrix is scaled in such way so that
$\tr(\Sigmab_1)=\tr(\Sigmab_2)$. We use $d=100$, $s=10$ and
$\epsilon=10^{-2}$. The prior precision matrix is
$\A=10^{-2}\,\I$. Figure \ref{f:deff} plots the scaled effective
dimension $d_{\frac nk\!\A}$ as a function of $k$, against the
full effective dimension for both examples. Unsurprisingly, for
the identity covariance the full effective dimension is almost
$d$, and the scaled effective dimension goes up very quickly to
match it. On the other hand, for the approximately low-rank
covariance, $d_{\A}\approx 55$ is considerably less then
$d=100$. Interestingly, the gap between the $d_{\frac nk\!\A}$ and
$d_{\A}$ for moderately small values of $k$ is even bigger. Our
theory suggests that $d_{\frac nk\!\A}$ is a valid indicator of
Bayesian degrees of freedom when $k\geq C\cdot d_{\frac nk\!\A}$
for some small constant $C$ (Theorem \ref{t:q1} has $C=4$, but we
believe this can be improved to $1$). While for the identity
covariance the condition $k\geq
  d_{\frac nk\!\A}$ is almost equivalent to $k\geq d_{\A}$, in the
approximately low-rank case, $k\geq d_{\frac nk\!\A}$ holds for
$k$ as small as 20, much less than $d_{\A}$.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{design/figs/deff_identity}\nobreak\includegraphics[width=0.5\textwidth]{design/figs/deff_lowrank}
  \caption{Scaled effective dimension compared to the full effective
  dimension for two diagonal data covariance matrices, with
  $\A=10^{-2}\,\I$.}
  \label{f:deff}
\end{figure}

@@

Given a psd matrix $\A$ and a weight vector $p=(p_1,...,p_n)\in[0,1]^n$, we
define $\DPPreg{p}(\X,\A)$ as a distribution over subsets $S\subseteq[n]$ (of all
sizes) such that (see Definition~\ref{d:r-dpp}):
\begin{align*}
  \Pr(S)\propto \det(
  \X_S^\top\X_S+\A)\ \cdot \prod_{i\in S}p_{i}\cdot\prod_{i\not\in
    S}(1-p_i).
\end{align*}
A number of regularized DPPs have been proposed recently
\citep{dpp-intermediate,regularized-volume-sampling}, mostly within the context
of Randomized Numerical Linear Algebra (RandNLA)~
\citep{Mah-mat-rev_JRNL,DM16_CACM,RandNLA_PCMIchapter_TR}.
To our knowledge, ours is the first such definition that strictly falls
under the umbrella of traditional DPPs \citep{dpp-ml}.
We show this in Section \ref{s:r-dpp}, where we also prove that
regularized DPPs can be decomposed into a low-rank DPP plus
i.i.d. Bernoulli sampling (Theorem \ref{t:algorithm}). This decomposition reduces
the sampling cost from $O(n^3)$ to $O(nd^2)$,
and involves a more general result about DPPs defined via a
correlation kernel (Lemma~\ref{l:decomp_corr_dpp}), which is of
independent interest.

@@

Purely greedy approximation algorithms have been shown to provide
guarantees in a number of special cases for experimental design. One
example is classical D-optimality criterion, which can be converted to
a submodular function \citep{submodularity-optimal-design}. Also, greedy algorithms
for Bayesian A/V-optimality criteria have been considered by
\cite{greedy-supermodular} and \cite{greedy-graph-sampling}. These
methods can only provide a constant factor approximation guarantee
(as opposed to $1+\epsilon$), and the factor is generally problem
dependent (which means it could be arbitrarily large).
Finally, a number of
heuristics with good empirical performance have been proposed, such as Fedorov's
exchange method \citep{cook1980comparison}. However, in this work
we focus on methods that provide theoretical approximation guarantees.
% In Section \ref{s:bayesian:experiments} we experimentally show that
% our regularized DPP sampling (without solving an SDP) performs
% better than iid sampling methods like uniform and predictive length
% sampling \cite{zhu2015optimal} and that by using the SDP solution
% we can improve it to match the performance of
% greedy algorithms for A-optimal design \cite{chamon2017approximate,greedy-supermodular}
% which empirically have been found to perform well on real world data.

\section{A new regularized determinantal point process}
\label{s:r-dpp}
In this section we develop the theory for a novel regularized extension of
determinantal point processes (DPP) which we use as the sampling distribution
for obtaining guarantees in Bayesian experimental design.
DPPs form a family of distributions which are used to model repulsion between
elements in a random set, with many applications in machine learning
\citep{dpp-ml}. Here, we focus on the setting where we are sampling out of all
$2^n$ subsets $S\subseteq[n]$. Traditionally, a DPP is defined by a
correlation kernel, which is an $n\times n$ psd matrix $\K$ with eigenvalues
between 0 and 1, i.e., such that $\zero\preceq\K\preceq \I$. Given a
correlation kernel $\K$, the corresponding DPP is
defined as
\begin{align*}
  S\sim\DPPcor(\K)\quad\text{iff}\quad
  \Pr(T\subseteq S) = \det(\K_{T,T})\ \  \forall_{T\in[n]},
\end{align*}
where $\K_{T,T}$ is the submatrix of $\K$ with rows and columns
indexed by $T$. Another way of defining a DPP, popular in the machine
learning community, is via an ensemble kernel $\L$. Any psd matrix
$\L$ is an ensemble kernel of a DPP defined as:
\begin{align*}
  S\sim\DPPens(\L)\quad\text{iff}\quad \Pr(S) \propto \det(\L_{S,S}).\quad~
\end{align*}
Crucially, every $\DPPens$ is also a $\DPPcor$, but not the other way
around. Specifically, $\DPPens(\L)=\DPPcor(\K)$ when:
\begin{align*}
   & \text{(a)}\ \  \L=\K(\I-\K)^{-1},\qquad
   & \text{(b)}\ \  \K = \I -  (\I+\L)^{-1},
\end{align*}
but (a) requires that $\I-\K$ be invertible which is not true
for some DPPs.
(This will be important in our analysis.)
The classical algorithm for sampling from a DPP requires the
eigendecomposition of either matrix $\K$ or $\L$, which in general
costs $O(n^3)$, followed by a sampling procedure which costs
$O(n\,|S|^2)$ \cite{dpp-independence,dpp-ml}.

We now define our regularized DPP and describe its
connection with correlation and ensemble~DPPs.
\begin{definition}\label{d:r-dpp}
  Given matrix $\X\in\R^{n\times d}$, a sequence  $p=(p_1,\dots,p_n)\in[0,1]^n$
  and a psd matrix $\A\in\R^{d\times d}$ such that
  $\sum_ip_i\x_i\x_i^\top\!+\A$ is
  full rank, let
  $\DPPreg{p}(\X,\A)$ be a distribution over $S\subseteq [n]$:
  \begin{align}
    \Pr(S) = \frac{\det(
      \X_S^\top\X_S+\A)}{\det\!\big(\!
      \sum_ip_i\x_i\x_i^\top\!+ \A\big)}
    \cdot \!\prod_{i\in S}p_{i}\cdot\!\prod_{i\not\in S}(1\!-\!p_i).\label{eq:poisson-prob}
  \end{align}
  %   \Pr(\pi)\ \propto\ \det\!\big(\A +
  %   \X_{\pi}^\top\X_{\pi}\big)
  %   \Pr\!\bigg(\Big[K\!=\!|\pi|\Big]\,\wedge\,\Big[\forall_{i=1}^{|\pi|}\,\pi_i\!=\!\pi_i\Big] \bigg).
  % \end{align*}
\end{definition}
The fact that this is a proper distribution (i.e., that it sums to
one) can be restated as a determinantal expectation formula: if
$b_i\sim\mathrm{Bernoulli}(p_i)$ are independent Bernoulli random
variables, then
\begin{align*}
   & \sum_{S\subseteq [n]}\!\det(
  \X_S^\top\X_S+\A)\prod_{i\in S}p_{i}\prod_{i\not\in S}(1-p_i) \\
   & \!=
  \E\bigg[\!\det\!\Big(\sum_ib_i\x_i\x_i^\top\!+\A\Big)\bigg]\overset{(*)}{=}
  \det\!\Big(\sum_i\E[b_i]\x_i\x_i^\top\!+\A\Big),
\end{align*}
where $(*)$ follows  from Lemma 7 of \citet{determinantal-averaging}.
% follows from a recently shown determinantal expectation formula.
% \begin{lemma}[\cite{determinantal-averaging}]\label{l:cb}
% For $\X\!\in\!\R^{n\times d}$, $\A\!\in\!\R^{d\times d}$
% and a random set $S\subseteq [n]$ drawn with independent Bernoullis,
% \begin{align*}
%   \E\Big[\det\!\big(\X_S^\top\X_S+\A\big)\Big]
%   = \det\!\Big( \E\big[\X_S^\top\X_S\big]+\A\Big).
% \end{align*}
% \end{lemma}

The main theoretical contribution in this section is the
following efficient algorithm for $\DPPreg{p}(\X,\A)$ which reduces it to
sampling from a correlation DPP and unioning with i.i.d.~Bernoulli samples:
\begin{theorem}\label{t:algorithm}
  For any $\X\in\R^{n\times d}$, $p\in[0,1]^n$ and a psd matrix $\A$
  s.t.~$\sum_ip_i\x_i\x_i^\top\!+\A$ is
  full rank, let
  \begin{align*}
    T & \sim\DPPcor\big(\D_p^{\sfrac12}\X(\A+\X^\top\D_p\X)^{-1}\X^\top\D_p^{\sfrac12}\big),
    \quad                                                                                    \\
      & \quad\text{where}\quad\D_p=\diag(p).
  \end{align*}
  If $b_i\sim\mathrm{Bernoulli}(p_i)$ are independent random variables, then
  $T\cup \{i:b_i\!=\!1\}\sim \DPPreg{p}(\X,\A)$.
\end{theorem}
